Sex-specific resilience of neocortex to food restriction

Version 1

Abstract

Abstract

Mammals have evolved sex-specific adaptations to reduce energy usage in times of food scarcity. These adaptations are well described for peripheral tissue, though much less is known about how the energy-expensive brain adapts to food restriction, and how such adaptations differ across the sexes. Here, we examined how food restriction impacts energy usage and function in the primary visual cortex (V1) of adult male and female mice. Molecular analysis and RNA sequencing in V1 revealed that in males, but not in females, food restriction significantly modulated canonical, energy-regulating pathways, including pathways associated with AMP-activated protein kinase (AMPK), peroxisome proliferator-activated receptor alpha (PPARα), mammalian target of rapamycin (mTOR), and oxidative phosphorylation. Moreover, we found that in contrast to males, food restriction in females did not significantly affect V1 ATP usage or visual coding precision (assessed by orientation selectivity). Decreased serum leptin is known to be necessary for triggering energy-saving changes in V1 during food restriction. Consistent with this, we found significantly decreased serum leptin in food-restricted males but no significant change in food-restricted females. Collectively, our findings demonstrate that cortical function and energy usage in female mice are more resilient to food restriction than in males. The neocortex, therefore, contributes to sex-specific, energy-saving adaptations in response to metabolic challenge.

Introduction

Mammals reduce their energy usage in times of food scarcity. These adaptations have been well documented for peripheral tissues and are sex-specific. For example, females, as compared to males, readily lose muscle and bone mass to reduce peripheral energy expenditure during food restriction, but lose less fat mass and overall bodyweight1–6. Moreover, females are more likely to suppress energy-costly, reproductive functions during food restriction than males, evidenced by substantive reductions in uterine and ovarian mass, as well as a cessation of reproductive function and behaviour7–9. Such sex-specific differences are thought to reflect the differential importance of different organs to sex-specific survival2.

In contrast to peripheral tissue, whether and how the brain adapts its function and energy use in a sex-specific manner during food scarcity remains largely unknown. The brain requires substantial amounts of energy to process and encode information10–12. Indeed, the human brain, which comprises 2% of our body’s mass, consumes 20% of our caloric intake, over half of which is used by the cerebral cortex12. Previously, we demonstrated that food restriction in male mice reduced energy usage in the primary visual cortex (V1) during visual processing, which was associated with a loss of orientation selectivity and visual coding precision13. These changes required a decrease in levels of serum leptin, a hormone secreted by adipocytes in proportion to fat mass14. Thus, in males the cortex, like peripheral tissue, has the capacity to reduce its energy usage and function in times of food scarcity.

By contrast, it remains unclear to what extent the brain similarly contributes to energy-saving adaptations during food scarcity in mammalian females. Understanding how food restriction impacts cortical function in males and females is not only of fundamental importance for understanding the sex-specific impact of diet on brain function, but is also critical for studies of cortical function, in which food restriction is used to behaviourally motivate animals15–17.

Here, we examined how food restriction leading to a 15% loss in body weight over 2-3 weeks impacts energy usage and visual coding in the primary visual cortex (V1) in adult male and female mice. We found that leptin levels, which regulate energy-saving changes in13, were maintained in females during food restriction, despite being decreased by approximately threefold in males. Consistent with these results, and using a range of experimental techniques, we failed to find significant energy-saving changes in V1 during food restriction in female mice, in contrast to male mice. Specifically, molecular analysis and RNA sequencing of V1 revealed that food-restriction significantly modified canonical energy-regulating cellular pathways in males, but not in females, including pathways associated with AMP-activated protein kinase (AMPK), peroxisome proliferator-activated receptor alpha (PPARα), and mammalian target of rapamycin (mTOR) signalling, as well as oxidative phosphorylation. Moreover, using in vivo ATP and calcium imaging, we found that food restriction significantly reduced V1 ATP use and orientation selectivity in males, but not in females.

Collectively, our study demonstrates that in times of food restriction, visual cortical function and energy usage are largely maintained in female mice, while they are reduced in males. The neocortex, therefore, contributes to sex-specific, energy-saving adaptations in response to metabolic stress. Our research highlights the importance of taking into account sex differences when studying the impact of dietary manipulations on brain function.

Results

We first examined sex-specific differences in how the neocortex adapts to food restriction in mice. Mice, 7-9 weeks of age, either had ad libwitum (CTR) access to food or were food restricted (FR) to 85% of their free feeding bodyweight, for 2-3 weeks (Figure 1A). Prior to any of our recordings, tissue collection, and analysis, all animals were given ad libitum access to food until sated. This enabled us to examine the long-term impact of caloric restriction on neocortical function, as opposed to the short-term impacts of hunger18,19.

Sex-specific impact of food restriction on weight loss and serum leptin.

(A) Animal weight across time. (B) Daily food intake (two-way ANOVA: CTR male vs. CTR female; p = 0.99; FR male vs. FR female; p = 0.07; n = 8 CTR males, 8 FR males, 7 CTR females, and 7 FR females). Percent reduction of food intake for food restriction is shown in red for each sex. (C) Animal weight (two-way ANOVA: n = 17 CTR males and 19 FR males; 22 CTR females and 11 FR females). (D) Serum leptin levels (two-way ANOVA: CTR females vs FR females; p=0.32; n = 17 CTR males and 19 FR males; 22 CTR females and 11 FR females). (E) Left. AMPK Thr172 phosphorylation, normalized by total AMPK, in primary visual cortex (V1) tissue (two-way ANOVA: CTR female vs. FR female; p = 0.11; n = 11 CTR males, 15 FR males, 11 CTR females, and 6 FR females). Right. PPARα activity in V1 tissue, as assessed by levels of DNA binding, normalized to protein level (two-way ANOVA: CTR female vs. FR female; p = 0.99; n = 5 CTR males, 11 FR males, n = 9 CTR females, and 9 FR females). *p < 0.05; n.s. = not significant. Error bars are S.E.M.

Sex-specific impact of food restriction on weight loss and fat mass-regulated hormone leptin

Prior to food restriction, we found that both males and females consumed the same ad libitum daily food intake, despite females weighing approximately 80% of age-matched males (weight: male vs. female; 29.42±0.20g vs 24.23±0.54g; t-test: p < 0.001; n = 8 male mice and 7 female mice). Critically, we found that to achieve a similar magnitude (15%) and rate of weight loss in both sexes we had to impose a 40% greater restriction of daily food intake in females (average of 39% of ad libitum intake; n = 7 female mice) compared to males (27% of ad libitum intake; n = 8 male mice) (Figure 1B,C). This is consistent with previous studies demonstrating that female mice are more resistant to weight loss than male1–6. Moreover, for similar bodyweight loss (15%) maintained over 2-3 weeks, serum leptin levels, which reflect fat mass, were markedly and significantly decreased (-72%) in males (t-test: p < 0.001; n = 17 control male mice and n = 19 food-restricted male mice), but only modestly and non-significantly decreased (-28%) in females (t-test: p = 0.32; n = 23 control female mice and n = 11 food-restricted female mice) (Figure 1D). This is in keeping with previous findings that female mice are more resistant to fat loss than males1–6 and that weight loss arises from other tissues including those associated with reproductive functions (ovarian and uterine mass)7–9.

Cellular energy-regulating pathways in cortex are significantly impacted by food restriction in males but not in females

Decreased leptin levels are necessary for triggering energy-saving changes in the mouse visual cortex during food restriction13. We therefore asked whether the absence of significant change in serum leptin in food-restricted females was associated with any change in V1 energy usage. We first assessed this at the molecular level by examining AMPK phosphorylation state. AMPK is a canonical cellular energy sensor activated by metabolic stress - such as food restriction, in part via decreased leptin signalling - through phosphorylation of threonine 172 (Thr172)20–22. Its activation results in reduced ATP expenditure, for example, through the inhibition of mammalian target of rapamycin (mTOR) signalling20,23, and regulation of mitochondrial oxidative phosphorylation24,25. In V1, we found that AMPK Thr172 phosphorylation was markedly and significantly elevated (2.9 fold) with food restriction in males (p < 0.05; n = 11 control male mice and n = 15 food-restricted male mice), consistent with ATP savings, but only modestly increased (1.4 fold) in females; this increase was not statistically significant (p = 0.11; n = 11 control female mice and n = 6 food-restricted female mice) (Figure 1E). We also examined the activity of peroxisome proliferator-activated receptor alpha (PPARα), which is downstream of leptin signalling and is known to regulate fatty acid metabolism and energy homeostasis22,26. As with AMPK, we found that PPARα was significantly regulated by food restriction selectively in males, but not in females (Figure 1E)27,28.

To further investigate how energy-regulating cellular pathways are impacted by food restriction, we performed RNA sequencing on tissue dissected from V1 (Figure 2A; n = 4 control male, n = 4 food-restricted male, n = 4 control female, and n = 4 food-restricted female mice). Differential gene expression analysis using DESeq2 revealed that food restriction significantly altered the expression of 657 and 585 targets in males and females, respectively (at a threshold of p-adj < 0.1; only 125 of these were common to both sexes (Figure 2B). To assess the functional relevance of the transcriptomic changes induced by food restriction, we performed Gene Set Enrichment Analysis (GSEA) to identify pathways enriched in the differentially expressed populations. We found that food restriction resulted in a significant alteration of 34 and 14 gene sets (at p-adj < 0.05) in males and females respectively; 13 of these were common to both sexes (Figure 2C; Table 1; Table 2). In male mice, we specifically found that food restriction significantly regulated gene sets associated with canonical energy-regulating pathways (Figure 2), including those associated with oxidative phosphorylation and mTOR signalling (mTORC1 signalling and PI3K/AKT/mTOR signalling), as well as fatty acid metabolism; these pathways are known to regulated by leptin and AMPK signaling20,27–29 (Figure 2D). Notably, these pathways were not significantly regulated by food-restriction in females. In addition to these findings, we found that gene sets associated to estrogen-regulating pathways were significantly regulated in food-restricted males and not in females (Figure 2C). This is consistent with previous findings showing that estrogen, like leptin, suppresses food intake and promotes energy expenditure30. Finally, consistent with past studies in cortex, we found that food restriction in males altered gene expression associated with inflammation and reactive oxidative species (Figure 2D)31,32.

Cellular energy-regulating pathways are more robustly impacted by food restriction in males than in females.

(A) Schema of RNA sequencing of V1 tissue. (B) Volcano plot of fold change expression with food restriction vs. significance for all analyzed genes; broken horizontal line denotes adjusted p < 0.1 significance level, below which non-significantly regulated expressed genes are marked by grey. Inset: Venn diagram showing the number of significantly differentially expressed genes with food restriction in males and females. (C) Gene Set Enrichment Analysis using Hallmark Gene Sets, depicted using a volcano plot of Enrichment Score vs. significance for males (left) and females (right); broken horizontal lines denotes adjusted p < 0.05 significance level. Inset: Venn diagram showing the number of significantly regulated gene sets with food restriction in males and females. (D) Complete list of significantly enriched gene sets by food-restriction in males and females for analysis in C. Gene sets of interest are denoted by a box. All data from 4 CTR males, 4 FR males, 4 CTR females, and 4 FR females.

Collectively these findings reveal that food restriction strongly affects energy-regulating cellular pathways in the cortex of male, as opposed to female, mice.

ATP usage in V1 is significantly decreased under food restriction in males but not in females

We next examined energy usage in visual cortex during visual processing using two photon ATP imaging in V1 of awake, head-fixed Thy1-ATeam1.03YEMK mice, which express a FRET-based ATP sensor33,34 (Figure 3A). In the presence of ATP synthesis inhibitors, focally applied over the visual cortex, ATP usage can be seen as a decay of the FRET signal during visual stimulation with natural stimuli13 (Figure 3A). Data obtained in females were compared to our previously published datasets obtained under similar conditions in male mice13. As previously shown13, food restriction robustly decreased the rate of cortical ATP usage in males (24% decrease) (t-test: p < 0.01; n = 11 control male mice and n = 10 food-restricted male mice); a more modest trend was observed in females (12% decrease), which was not significant compared to controls (t-test: p = 0.18; n = 12 control female mice and n = 8 food-restricted female mice) (Figure 3B-C). These findings demonstrate that food-restriction significantly impacts V1 energy usage in males but not in females.

ATP use and orientation selectivity in V1 is more robustly decreased by food restriction in males.

(A) Schemata of ATP imaging experiment. Top: example field of view of V1 layer 2/3 neurons in the ATeam1.03YEMK transgenic mouse (scale bar: 10 μm). Bottom: ATP synthesis inhibitors were used to isolate ATP usage, recorded as a decrease in FRET signal during presentation of natural, outdoor scenes. (B) Normalized ATeam1.03YEMK FRET signal in males and during presentation of natural, outdoor scenes. ATP synthesis inhibitors (arrow) were added to isolate ATP usage. Inset: Time to 50% decay of the ATeam1.03YEMK FRET signal (two-way ANOVA; n = 11 CTR males and 10 FR males). (C) As in (B) but for females (inset: two-way ANOVA; p = 0.18; n = 12 CTR females and n = 8 FR females). (D) Top: Schema of two-photon imaging and sample field of view of V1 layer 2/3 neurons expressing GCaMP6s (scale bar: 10 μm). Bottom: Sample fluorescent signals (grey) from a selected neuron (black circle) in response to 6 drifting gratings of varying orientations. Trial averages are in black. Horizontal bar denotes 2s grating presentation. (E) Mean orientation tuning normalized to the response to the preferred orientation for CTR and FR males (data from 13). Note that -90° and +90° conditions correspond to the same visual stimulus. Inset: Orientation selectivity index (OSI; two-way ANOVA; n = 8 CTR and 8 FR males). (F) As in (E) but for females (Inset: two-way ANOVA; p = 0.08; n = 7 CTR females and 9 FR females). **p < 0.01; ***p < 0.001; n.s. = not significant. Error bars are S.E.M. Data from male mice are from a previously published dataset obtained under similar conditions13.

Orientation selectivity in V1 is significantly reduced under food restriction in males but not in females

We next examined how food restriction impacts visual cortical function. In male mice, reductions in energy usage in V1 during food restriction are associated with a decreased visual coding precision as evidenced by a reduction of orientation selectivity13. Given the non-significant impact of food-restriction on energy usage in female mice, we reasoned that orientation selectivity would also be less impacted. We analysed two-photon calcium imaging data of V1 GCaMP6s-labeled layer 2/3 neurons in female mice viewing drifting gratings (Figure 3D-F; n = 7 control male, n = 8 food-restricted male, n = 7 control female, and n = 9 food-restricted female mice). Data obtained in females were compared to our previously published datasets obtained under similar conditions in male mice13. Consistent with our hypothesis, in contrast to males in which orientation selectivity was significantly reduced by 27% (p<0.001), orientation selectivity was modestly and not significantly reduced by food-restriction in females (13%; p=0.08) (Figure 3E-F).

Altogether, our results show that energy usage and coding precision in the visual cortex are largely maintained in female mice under food restriction while they are robustly reduced in males.

Bayes factor analysis reveals statistically robust impact of metabolic stress on V1 energy use and visual coding in males, but not females

We used Bayes factor hypothesis testing to better assess the statistical impact of food restriction on visual cortical function and energy use in male and female mice (Supplemental Figure 1). Bayes factor analysis uses a Bayesian framework to quantify how much more likely the data is under the alternative hypothesis (i.e. there is an effect of food restriction) versus the null hypothesis (i.e. there is no effect of food restriction), quantified by the Bayes factor (BF10)35,36. A BF10 > 3, for example, means that the data is >3 times more likely under the alternative hypothesis than the null hypothesis. Critically, the BF10 allows for a more informative comparison of the effects of food restriction between sexes, as opposed to comparing p-values, which can be misleading35,36. Across the parameters of energy use and visual coding that we tested (Figures 1-3), BF10 for males was consistently several-fold greater than that for females. Thus, visual cortical function and energy use was more robustly modulated by food restriction in males than in females.

Our findings however cannot rule out a potential, smaller impact of food restriction on cortical function and energy use in females, below the threshold for statistical detection in this study. The Bayes factor analysis helps to differentiate the absence of evidence from the evidence of absence. In particular, moderate evidence in support for the absence of an effect requires a BF10 < 0.33 (i.e. the data is 3 times more likely under the null hypothesis than the alternative hypothesis)35,36. For females, we found only 3 out of the 10 parameters had BF10 < 0.33 (Supplemental Figure 1), suggesting that food restriction may have a modest impact on V1 energy use or function, but below the limits of statistical detection in our study. Altogether, our results show that energy usage and coding precision in the visual cortex are largely maintained in female mice under food restriction while they are robustly reduced in males.

Discussion

Here, we find that in female mice, energy usage and visual coding in V1 is largely unchanged under food restriction. Using molecular analysis and RNA sequencing, we found that energy-regulating pathways were not significantly changed with food restriction in females, in contrast to males; these included pathways associated with AMPK, mTOR, and PPARα signalling, as well as oxidative phosphorylation. Consistent with this, we found that food restriction in females did not impact visual cortical ATP usage nor orientation selectivity, both of which were reduced by food restriction in males. Our findings suggest that the contribution of neocortex to energy-saving adaptations during food restriction is sex-selective and more pronounced in males; females may rely on other energy-saving strategies, such as reducing reproductive function, to cope with food restriction.

Limitations of the study

Our study focused on the effects of a moderate food restriction protocol over 2-3 weeks on neocortical function and energy use. This protocol is widely used in metabolic studies investigating the impact of caloric restriction on peripheral physiology, as well as in neuroscience studies to motivate rodents to perform specific behaviors6,15–17. However, it remains unclear to what extent our findings generalize to other protocols of food restriction. Indeed, the duration, magnitude, and pattern of food restriction, as well as macronutrient composition are known to impact the physiological responses to dietary manipulations22,37–39. Moreover, whilst we focussed our study on the adult visual cortex, it remains unclear if the brain shows similar sex-specific energy-saving adaptations in other cortical areas, and how such adaptations change with age6,40.

A key strength of our study is that we consistently found, across a number of experimental methods, that changes in cortical function and energy use were significant in males, but not in females. However, a critical limitation of our study is that we did not have the statistical power to directly compare the effect size of food restriction between the sexes (e.g. ANOVA interaction effects); to do so would require a substantial increase in experimental group sizes by several fold. We addressed this limitation, in part, using Bayes factor analysis, which revealed that the impact of food restriction in males was indeed more robust than in females. This analysis also revealed that our study did not have sufficiently strong evidence to definitively rule out a smaller effect of food restriction in females, which may not have been statistically detectable.

Role of leptin and sex hormones in food restriction impact on cortical activity

Consistent with past studies, we found that food-restricted female mice resisted weight loss, and therefore required a greater restriction of food to achieve similar weight loss (relative to baseline) as males. In addition, for the same relative weight loss as males, females exhibited minimal reductions in serum leptin levels (secreted by adipocytes), consistent with past studies showing that females resist fat loss during food restriction1–3,5,6. Similar effects have been observed across several mammalian species, including humans, and is thought to reflect the differential importance of fat stores for reproductive success in males and females1–5. Notably, we previously demonstrated that reductions in leptin levels during food restriction are necessary for driving energy-saving changes in visual cortical coding in male mice; leptin therefore acts as a critical signal linking the neocortex to fat stores13. That leptin levels are not significantly affected by food restriction in females may explain the minimal impact food restriction has on cortical energy usage and coding precision in female mice.

In addition to leptin, sex hormones likely contribute to sex-specific adaptations to food restriction41. As with leptin, estrogen suppresses food intake and promotes energy expenditure30, in part by augmenting leptin signalling by increasing leptin-induced phosphorylation of STAT342. Moreover, estrogen can enhance ATP production by sensitizing insulin signalling, enhancing glycolysis and oxidative phosphorylation, and along with progesterone, upregulating mitochondrial gene expression and function30,43,44. Interestingly, in our study, RNA sequencing revealed that estrogen signalling pathways were significantly regulated in male visual cortex with food-restriction, and were unchanged in females. Estrogen may thus have overlapping effects with those induced by food restriction, thereby diminishing further responses to food restriction in females6. Notably, it was recently shown that sex differences in calorie-restriction’s metabolic effects that are observed in young adult mice, are largely absent in older mice, when females’ estrogen levels have declined6. Further studies will be needed to explore whether the resilience of cortical function found in this study in young cycling adult females is also observed in non-cycling mice with reduced oestrogen levels, such as in aged mice.

Sex-specific regulation of cellular energy-regulating pathways

AMPK and mTOR signalling pathways are canonical regulators of cellular energy usage20,23. We found these pathways, along with cortical ATP usage, to be significantly modulated by food-restriction in male, but not female mice. AMPK is activated during metabolic stress, and saves energy by regulating oxidative phosphorylation24,25, and inhibiting cellular pathways associated with energy usage, such as mTOR signalling20. mTOR signalling, by upregulating protein synthesis, increases ATP use by promoting neuronal excitability, neurite outgrowth, ion channel expression, and synaptic plasticity23. mTOR signalling is therefore ideally situated to regulate how much ATP is used in neuronal function; its inhibition may underlie ATP savings and the associated loss of coding precision with food-restriction in male mice. Notably, previous studies have found sex- and tissue-dependent regulation of AMPK and mTOR signalling, including in response to food restriction41. For example, in skeletal muscle, mTOR signalling is enhanced by fasting selectively in male, but not female mice41,45. Moreover, metabolic challenge by endurance training induces greater AMPK activity in cardiac tissue in male, as opposed to female mice46. Our study extends the sex-specific regulation of these energy-regulating pathways to the neocortex.

Importance of biological sex on the impact of food restriction on cortical function

Food restriction protocols are widely used to motivate animals in behavioural studies examining cortical function15–17. Our findings suggest that these protocols are likely to have greater impacts on basal neocortical function in males, as compared to females. Consistent with this, a recent study has found that food restriction in juvenile mice strongly impacts cognitive flexibility in learning and decision-making in adulthood, specifically for males but not females47,48. Such sex-specific impacts of food restriction on neocortical function are important to take into consideration in the design and interpretation of such studies.

Diet is known to have a considerable impact on neurological diseases, including epilepsy, Alzheimer’s disease, and other neurodegenerative diseases49,50. Whilst calorie restricted diets have proven to be of benefit in this regard, the sex-dependency of such interventions have not been thoroughly investigated, though on the basis of our findings it is conceivable that males and females may respond differently to dietary interventions. The inclusion of both males and females in biomedical research, and the study of sex differences, is therefore of considerable importance for ensuring efficacious treatments are developed for both sexes.

In conclusion, we find that visual cortical coding and energy usage in females are resistant to food restriction, in contrast to males. Our findings establish that sex-specific adaptations in energy usage, previously described in peripheral tissue under metabolic challenge, extends to the neocortex.

eLife Assessment

This study provides important findings based on convincing evidence demonstrating that females and males have different strategies to regulate energy consumption in the brain in the context of low energy intake. While food deprivation reduces energy consumption and visual processing performance in the visual cortex of males, the female cortex is unaffected, likely at the expense of other functions. This study is relevant for scientists interested in body metabolism and neuroscience.

Reviewer #1 (Public Review):

Padamsey et al. followed up on their previous study in which they found that male mice sacrifice visual cortex computation precision to save energy in periods of food restriction (Padamsey et al. 2021, Neuron). In the present study, the authors find that female mice show much lower levels of adaptation in response to food restriction on the level of metabolic signaling and visual cortex computation. This is an important finding for understanding sex differences in adaptation to food scarcity and also impacts the interpretation of studies employing food restriction in behavioral analyses and learning paradigms.

The manuscript is, in general, very clear and the conclusions are straightforward. The main limitation, that the number of experiments is insufficient to compare the effects of food restriction in males and females directly, is discussed by the authors: to address this point they use Bayes factor analysis to provide an estimate of the likelihood that females and males indeed differ in terms of energy metabolism and sensory processing adaptions during food restriction.

The following points are not entirely clear yet.

1. For a number of experiments the authors use their new data set on females and compare that with the data set previously published on males. In how far are these data sets comparable? Have they been performed originally in parallel for example using siblings of different sexes or have the experiments been conducted several years apart from each other? What is the expected variability, if one repeated these experiments with the same sex considering the differences/similarities between experimental setups, housing conditions, interindividual differences, etc.?

2. Energy consumption and visual processing may differ between periods in which animals are in different behavioral states. Is there a possibility that male and female mice differed in behavioral state during measurements? Were animals running or resting during visual stimulation and during ATP measurements?

3. Related to the previous point: the authors show that ATP consumption was reduced in male mice during visual stimulation. What about visual cortex ATP consumption in the absence of visual stimulation? Do food-deprived males and/or females show lower ATP consumption in the visual cortex e.g. during sleep?

Reviewer #2 (Public Review):

Summary:

Padamsey et al build up on previous significant work from the same group which demonstrated robust changes in the visual cortex in male mice from long-term (2-3 weeks) food restriction. Here, the authors extend this finding and reveal striking sex-specific differences in the way the brain responds to food restriction. The measures included the whole-body measure of serum leptin levels, and V1-specific measures of activity of key molecular players (AMPK and PPARα), gene expression patterns, ATP usage in V1, and the sharpness of visual stimulus encoding (orientation tuning). All measures supported the conclusion that the female mouse brain (unlike in males) does not change its energy usage and cortical functional properties on comparable food restriction.

While the effect of food restriction on more peripheral tissue such as muscle and bones has been well studied, this result contributes to our understanding of how the brain responds to food restriction. This result is particularly significant given that the brain consumes a large fraction of the body's energy consumption (20%), with the cortex accounting for half of that amount. The sex-specific differences found here are also relevant for studies using food restriction to investigate cortical function.

Strengths:

The study uses a wide range of approaches mentioned above which converge on the same conclusion, strengthening the core claim of the study.

Weaknesses:

Since the absence of a significant effect does not prove the absence of any changes, the study cannot claim that the female mouse brain does not change in response to food restriction. However, the authors do not make this claim. Instead, they make the well-supported claim that there is a sex-specific difference in the response of V1 to food restriction.

Reviewer #3 (Public Review):

Summary:

The authors food-deprived male and female mice and observed a much stronger reduction of leptin levels, energy consumption in the visual cortex, and visual coding performance in males than females. This indicates a sex-specific strategy for the regulation of the energy budget in the face of low food availability.

Strengths:

This study extends a previous study demonstrating the effect of food deprivation on visual processing in males, by providing a set of clear experimental results, demonstrating the sex-specific difference. It also provides hypotheses about the strategy used by females to reduce energy budget based on the literature.

Weaknesses:

The authors do not provide evidence that females are not impacted by visually guided behaviors contrary to what was shown in males in the previous study.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1:

(1) For a number of experiments the authors use their new data set on females and compare that with the data set previously published on males. In how far are these data sets comparable? Have they been performed originally in parallel for example using siblings of different sexes or have the experiments been conducted several years apart from each other? What is the expected variability, if one repeated these experiments with the same sex considering the differences/similarities between experimental setups, housing conditions, interindividual differences, etc.?

This is an important point. We did our best to collect the data in similar conditions (same set-ups; same animal housing conditions) and in experimental cohorts including both males and females. While some data from males were published first, the acquisition of male and female data was done in the same time period.

Specifically, all results shown in Figure 1 and Figure 2 (Serum leptin, PPARalpha, AMPK, RNAseq) come from samples (from both males and females) that were processed at the same time and in similar conditions, by the same authors (Z.P. and P. M.).

For the in vivo data (Figure 3, Supplementary figure 1), the male and female data were collected within a 1–2-year timeframe, in the same setups, by the same two authors (Z.P., D.K.). The males and females were housed under similar conditions (same room, same cage type, in groups of 25). We did not use siblings of different sexes. Independent cohorts (1-12 months apart), including both males and females, went into each data set. The within cohort variability does not obviously differ from between cohort variability, however the n number of animals is too small to confirm this with sufficient statistical power.

Altogether, the differences observed between male and female data cannot be explained by the timing and conditions of data acquisition from both sexes.

(2) Energy consumption and visual processing may differ between periods in which animals are in different behavioral states. Is there a possibility that male and female mice differed in behavioral state during measurements? Were animals running or resting during visual stimulation and during ATP measurements?

We thank the reviewer for this suggestion. We have now edited the text and included a new supplementary figure. All in vivo experiments were done in stationary animals that were resting in a cardboard tube both during 2-photon imaging and ATP measurements. Animals were also well habituated to the setup. In addition, we have imaged pupil diameters during in vivo imaging session. We have quantified pupil diameter during visual stimulation and do not find a sex difference (Supplemental Figure 2). Thus, we did not find a significant difference in behavioural or attentional state between sexes, in our experimental conditions.

We have edited the text to include this information (lines 183-185).

(3) Related to the previous point: the authors show that ATP consumption was reduced in male mice during visual stimulation. What about visual cortex ATP consumption in the absence of visual stimulation? Do food-deprived males and/or females show lower ATP consumption in the visual cortex e.g. during sleep?

We have repeated V1 ATP imaging experiments in the dark, in the absence of visual stimulation, in both males and females (Supplementary figure 1). ATP consumption rates are slower in the dark vs. during visual stimulation. Moreover, we find that in the dark, there is no difference in ATP consumption rate between control and food restricted animals of either sex. Thus, the reduced ATP consumption we found with food restriction in males is related specifically to the active processing of visual information.

We have edited the text to include this information (lines 158-159).

Reviewer 2:

(1) It appears that the authors have the data for doing decoding analysis, similar to Fig 6D in their previous paper. However, this analysis has not been done for this study. This would be good to include. If the authors have attempted the behavioural discrimination tests on female mice as in the previous study, this would also be useful to include.

The first point of the reviewer is about datasets acquired in males that are included in our previous publication (Padamsey et al., 2022) but not compared to female data in the present manuscript.

Whilst we fully agree that these results would be very useful, we did not have the resources (in terms of skilled researcher and funding) to perform these experiments in female mice. That is why these results are not included in this manuscript.

(2) There appears to be an inconsistency in the methods of reporting OSI. It states that the OSI of grating-responsive neurons was calculated as 1 - circular variance. But then OSI is defined as simply abs(). Also, it would be good to be consistent about reporting medians as the median without confounding with the average (which is the mean). Sentences such as the following do not make sense: The average OSI for an animal was taken as the median OSI value calculated across neurons. This should be corrected throughout the manuscript, where the average is mentioned but the median is measured.

We thank the reviewer for noting this issue and we apologize for the confusion. We have now clarified the above in the manuscript (lines 587-603) and insert the following reference for the detailed explanation of OSI and DSI calculation: Mazurek M, Kager M, Van Hooser SD. Robust quantification of orientation selectivity and direction selectivity. Front Neural Circuits. 2014. https://doi.org/10.3389/fncir.2014.00092

In the figure showing the orientation tuning, the authors have collapsed the two directions of each orientation together. However, if I understand correctly, the calculation of OSI does not do this step of collapsing. In this case, and in the interest of revealing more useful features of the data instead of averaging them out, it would be good to show the average tuning curves with and without FR for all directions, not collapsed.

As with orientation tuning, we found that direction tuning is reduced with food restriction, and that this is significant in males, but not in females. These results are now included in the text, with statistics (lines 179-180) and in Supplemental Figure 3.

Reviewer 3:

l. 183-187 The discussion based on the idea that "The Bayes factor analysis helps to differentiate the absence of evidence from the evidence of absence." does not seem very helpful. Using a statistical criterium makes less sense than providing the reader with an estimate largest effect size (if there is any) that is compatible with the observation. If there would be a significant effect but of a very small size would it change the authors' conclusion? That seems unlikely. I recommend removing the sentence on line 184, which is in fact not used afterwards.

We agree with the reviewer. We have now removed the sentence and rephrased (lines 202-208).

Editor's note:

Should you choose to revise your manuscript, please include full statistical reporting including exact pvalues wherever possible alongside the summary statistics (test statistic and df) and 95% confidence intervals. These should be reported for all key questions and not only when the p-value is less than 0.05.

We now provide exact p-values alongside the summary statistics (test statistic and df) and 95% confidence intervals for all key results.

Version 2

Abstract

Abstract

Mammals have evolved sex-specific adaptations to reduce energy usage in times of food scarcity. These adaptations are well described for peripheral tissue, though much less is known about how the energy-expensive brain adapts to food restriction, and how such adaptations differ across the sexes. Here, we examined how food restriction impacts energy usage and function in the primary visual cortex (V1) of adult male and female mice. Molecular analysis and RNA sequencing in V1 revealed that in males, but not in females, food restriction significantly modulated canonical, energy-regulating pathways, including pathways associated with AMP-activated protein kinase (AMPK), peroxisome proliferator-activated receptor alpha (PPARα), mammalian target of rapamycin (mTOR), and oxidative phosphorylation. Moreover, we found that in contrast to males, food restriction in females did not significantly affect V1 ATP usage or visual coding precision (assessed by orientation selectivity). Decreased serum leptin is known to be necessary for triggering energy-saving changes in V1 during food restriction. Consistent with this, we found significantly decreased serum leptin in food-restricted males but no significant change in food-restricted females. Collectively, our findings demonstrate that cortical function and energy usage in female mice are more resilient to food restriction than in males. The neocortex, therefore, contributes to sex-specific, energy-saving adaptations in response to metabolic challenge.

Introduction

Mammals reduce their energy usage in times of food scarcity. These adaptations have been well documented for peripheral tissues and are sex-specific. For example, females, as compared to males, readily lose muscle and bone mass to reduce peripheral energy expenditure during food restriction, but lose less fat mass and overall bodyweight1–6. Moreover, females are more likely to suppress energy-costly, reproductive functions during food restriction than males, evidenced by substantive reductions in uterine and ovarian mass, as well as a cessation of reproductive function and behaviour7–9. Such sex-specific differences are thought to reflect the differential importance of different organs to sex-specific survival2.

In contrast to peripheral tissue, whether and how the brain adapts its function and energy usage in a sex-specific manner during food scarcity remains largely unknown. The brain requires substantial amounts of energy to process and encode information10–12. Indeed, the human brain, which comprises 2% of our body’s mass, consumes 20% of our caloric intake, over half of which is used by the cerebral cortex12. Previously, we demonstrated that food restriction in male mice reduced energy usage in the primary visual cortex (V1) during visual processing, which was associated with a loss of orientation selectivity and visual coding precision13. These changes required a decrease in levels of serum leptin, a hormone secreted by adipocytes in proportion to fat mass14. Thus, in males the cortex, like peripheral tissue, has the capacity to reduce its energy usage and function in times of food scarcity.

By contrast, it remains unclear to what extent the brain similarly contributes to energy-saving adaptations during food scarcity in mammalian females. Understanding how food restriction impacts cortical function in males and females is not only of fundamental importance for understanding the sex-specific impact of diet on brain function, but is also critical for studies of cortical function, in which food restriction is used to behaviourally motivate animals15–17.

Here, we examined how food restriction leading to a 15% loss in body weight over 2-3 weeks impacts energy usage and visual coding in the primary visual cortex (V1) in adult male and female mice. We found that leptin levels, which regulate energy-saving changes in13, were maintained in females during food restriction, despite being decreased by approximately threefold in males. Consistent with these results, and using a range of experimental techniques, we failed to find significant energy-saving changes in V1 during food restriction in female mice, in contrast to male mice. Specifically, molecular analysis and RNA sequencing of V1 revealed that food-restriction significantly modified canonical energy-regulating cellular pathways in males, but not in females, including pathways associated with AMP-activated protein kinase (AMPK), peroxisome proliferator-activated receptor alpha (PPARα), and mammalian target of rapamycin (mTOR) signalling, as well as oxidative phosphorylation. Moreover, using in vivo ATP and calcium imaging, we found that food restriction significantly reduced V1 ATP usage and orientation selectivity in males, but not in females.

Collectively, our study demonstrates that in times of food restriction, visual cortical function and energy usage are largely maintained in female mice, while they are reduced in males. The neocortex, therefore, contributes to sex-specific, energy-saving adaptations in response to metabolic stress. Our research highlights the importance of taking into account sex differences when studying the impact of dietary manipulations on brain function.

Results

We first examined sex-specific differences in how the neocortex adapts to food restriction in mice. Mice, 7-9 weeks of age, either had ad libwitum (CTR) access to food or were food restricted (FR) to 85% of their free feeding bodyweight, for 2-3 weeks (Figure 1A). Prior to any of our recordings, tissue collection, and analysis, all animals were given ad libitum access to food until sated. This enabled us to examine the long-term impact of caloric restriction on neocortical function, as opposed to the short- term impacts of hunger18,19.

Sex-specific impact of food restriction on weight loss and serum leptin.

(A) Animal weight across time. (B) Daily food intake (two-way ANOVA: CTR male vs FR male; t = 4.81; df = 25; p < 0.0001; CTR male vs. CTR female; t = 0.013; df = 25; p = 0.99; CTR female vs. FR female; t = 6.59; df=25; p < 0. 0001; FR male vs. FR female; t = 1.93; df = 25; p = 0.07; n = 8 CTR males, 8 FR males, 7 CTR females, and 7 FR females). Percent reduction of food intake for food restriction is shown in red for each sex. (C) Animal weight (two-way ANOVA: CTR male vs FR male; t = 11.36; df = 25; p < 0.0001; CTR female vs FR female; t = 8.36; df = 25; p < 0.0001; n = 17 CTR males and 19 FR males; 23 CTR females and 11 FR females). (D) Serum leptin levels (two-way ANOVA: CTR male vs FR male; t = 4.58; df = 66; p < 0.0001; CTR females vs FR females; t = 1.00; df = 66; p = 0.32; n = 17 CTR males and 19 FR males; 23 CTR females and 11 FR females). (E) Left. AMPK Thr172 phosphorylation, normalized by total AMPK, in primary visual cortex (V1) tissue (two-way ANOVA: CTR male vs FR male; t = 2.28; df = 39; p = 0.022; CTR female vs. FR female; t = 0.64; df = 39; p = 0.11; n = 11 CTR males, 15 FR males, 11 CTR females, and 6 FR females). Right. PPARα activity in V1 tissue, as assessed by levels of DNA binding, normalized to protein level (two-way ANOVA: CTR male vs FR male; t = 4.81; df = 30; p = 0.013; CTR female vs. FR female; t = 0.0016; df = 30; p = 0.99; n = 5 CTR males, 11 FR males, n = 9 CTR females, and 9 FR females). ***p < 0.0001; *p < 0.05; n.s. = not significant. Error bars are S.E.M.

Sex-specific impact of food restriction on weight loss and fat mass-regulated hormone leptin

Prior to food restriction, we found that both males and females consumed the same ad libitum daily food intake, despite females weighing approximately 80% of age-matched males (weight: male vs. female; 27.82 g [95% CI: 26.84 g – 28.79 g] vs 23.21 g [95% CI: 22.35 g – 24.07 g]; t = 7.58; df = 66; p < 0.001; n = 8 male and 7 female mice). Critically, we found that to achieve a similar magnitude (15%) and rate of weight loss in both sexes we had to impose a 40% greater restriction of daily food intake in females (average of 39% of ad libitum intake; n = 7 mice) compared to males (27% of ad libitum intake; n = 8 mice) (Figure 1B,C). This is consistent with previous studies demonstrating that female mice are more resistant to weight loss than male1–6. Moreover, for similar bodyweight loss (15%) maintained over 2-3 weeks, serum leptin levels, which reflect fat mass, were markedly and significantly decreased (-72%) in males (leptin: CTR male vs FR male; 6.45 ng/mL [95% CI: 4.47 ng/mL – 8.43 ng/mL] vs. 1.84 ng/mL [95% CI: 0.90 ng/mL – 2.79 ng/mL]; t = 4.58; df = 66; p < 0.0001; n = 17

CTR males and 19 FR males), but only modestly and non-significantly decreased (-28%) in females (leptin: CTR females vs FR females; 3.94 ng/mL [95% CI: 2.53 ng/mL – 5.35 ng/mL] vs. 2.83 ng/mL [95% CI: 1.20 ng/mL – 4.47 ng/mL]; t = 1.00; df = 66; p = 0.32; 23 CTR females and 11 FR females) (Figure 1D). This is in keeping with previous findings that female mice are more resistant to fat loss than males1–6 and that weight loss arises from other tissues including those associated with reproductive functions (ovarian and uterine mass)7–9.

Cellular energy-regulating pathways in cortex are significantly impacted by food restriction in males but not in females

Decreased leptin levels are necessary for triggering energy-saving changes in the mouse visual cortex during food restriction13. We therefore asked whether the absence of significant change in serum leptin in food-restricted females was associated with any change in V1 energy usage. We first assessed this at the molecular level by examining AMPK phosphorylation state. AMPK is a canonical cellular energy sensor activated by metabolic stress - such as food restriction, in part via decreased leptin signalling - through phosphorylation of threonine 172 (Thr172)20–22. Its activation results in reduced ATP expenditure, for example, through the inhibition of mammalian target of rapamycin (mTOR) signalling20,23, and regulation of mitochondrial oxidative phosphorylation24,25. In V1, we found that AMPK Thr172 phosphorylation was markedly and significantly elevated (2.9 fold) with food restriction in males (AMPK Thr172 phosphorylation: CTR male vs FR male; 13.19 AU/μg [95% CI: 7.68 AU/μg – 18.71 AU/μg] vs. 38.21 AU/μg [95% CI: 16.50 AU/μg – 59.93 AU/μg]; t = 2.28; df = 39; p = 0.022; n = 11 CTR males and 15 FR males), consistent with ATP savings, but only modestly increased (1.4 fold) in females; this increase was not statistically significant (AMPK Thr172 phosphorylation: CTR female vs. FR female; 20.94 AU/μg [95% CI: 9.42 AU/μg - 32.46 AU/μg] vs. 29.52 AU/μg [95% CI: 8.52 AU/μg – 50. 53 AU/μg]; t = 0.64; df = 39; p = 0.11; 11 CTR females, and 6 FR females) (Figure 1E). We also examined the activity of peroxisome proliferator-activated receptor alpha (PPARα), which is downstream of leptin signalling and is known to regulate fatty acid metabolism and energy homeostasis22,26. As with AMPK, we found that PPARα was significantly regulated by food restriction selectively in males, but not in females (Figure 1E)27,28.

To further investigate how energy-regulating cellular pathways are impacted by food restriction, we performed RNA sequencing on tissue dissected from V1 (Figure 2A; n = 4 CTR male, n = 4 FR male, n = 4 CTR female, and n = 4 FR female mice). Differential gene expression analysis using DESeq2 revealed that food restriction significantly altered the expression of 657 and 585 targets in males and females, respectively (at a threshold of p-adj < 0.1; only 125 of these were common to both sexes (Figure 2B). To assess the functional relevance of the transcriptomic changes induced by food restriction, we performed Gene Set Enrichment Analysis (GSEA) to identify pathways enriched in the differentially expressed populations. We found that food restriction resulted in a significant alteration of 34 and 14 gene sets (at p-adj < 0.05) in males and females respectively; 13 of these were common to both sexes (Figure 2C; Table 1; Table 2). In male mice, we specifically found that food restriction significantly regulated gene sets associated with canonical energy-regulating pathways (Figure 2), including those associated with oxidative phosphorylation and mTOR signalling (mTORC1 signalling and PI3K/AKT/mTOR signalling), as well as fatty acid metabolism; these pathways are known to regulated by leptin and AMPK signaling20,27–29 (Figure 2D). Notably, these pathways were not significantly regulated by food-restriction in females. In addition to these findings, we found that gene sets associated to estrogen-regulating pathways were significantly regulated in food-restricted males and not in females (Figure 2C). This is consistent with previous findings showing that estrogen, like leptin, suppresses food intake and promotes energy expenditure30. Finally, consistent with past studies in cortex, we found that food restriction in males altered gene expression associated with inflammation and reactive oxidative species (Figure 2D)31,32.

Cellular energy-regulating pathways are more robustly impacted by food restriction in males than in females.

(A) Schema of RNA sequencing of V1 tissue. (B) Volcano plot of fold change expression with food restriction vs. significance for all analyzed genes; broken horizontal line denotes adjusted p < 0.1 significance level, below which non-significantly regulated expressed genes are marked by grey. Inset: Venn diagram showing the number of significantly differentially expressed genes with food restriction in males and females. (C) Gene Set Enrichment Analysis using Hallmark Gene Sets, depicted using a volcano plot of Enrichment Score vs. significance for males (left) and females (right); broken horizontal lines denotes adjusted p < 0.05 significance level. Inset: Venn diagram showing the number of significantly regulated gene sets with food restriction in males and females. (D) Complete list of significantly enriched gene sets by food-restriction in males and females for analysis in C. Gene sets of interest are denoted by a box (oxidative phosphorylation: male; t = 0.48; p < 0.0001; female; t = -0.17; p = 0.96; mTORC1 signaling: male; t = 0.46; p < 0.0001; female; t = -0.18; p = 0.98; fatty acid metabolism: male; t = 0.43; p < 0.0001; t = 0.2; p = 0.88; PI3K/AKT/mTOR signaling: male; t = 0.34; p = 0.027; female; t = 0.17; p = 0.98). All data from 4 CTR males, 4 FR males, 4 CTR females, and 4 FR females.

Collectively these findings reveal that food restriction strongly affects energy-regulating cellular pathways in the cortex of male, as opposed to female, mice.

ATP usage in V1 is significantly decreased under food restriction in males but not in females

We next examined energy usage in visual cortex during visual processing using two photon ATP imaging in V1 of awake, head-fixed Thy1-ATeam1.03YEMK mice, which express a FRET-based ATP sensor33,34 (Figure 3A). In the presence of ATP synthesis inhibitors, focally applied over the visual cortex, ATP usage can be seen as a decay of the FRET signal during visual stimulation with natural stimuli13 (Figure 3A). Data obtained in females were compared to our previously published datasets obtained under similar conditions in male mice13. As previously shown13, food restriction robustly decreased the rate of cortical ATP usage in males (24% decrease) (half-time of FRET decay: CTR male vs. FR male; 8.48 min [95% CI: 7.59 min – 9.37 min] vs. 10.51 min [95% CI: 9.08 min – 11.93 min]; t = 2.87; df = 37; p = 0. 0067; n = 11 CTR males and 10 FR males) during visual stimulation; a more modest trend was observed in females (12% decrease), which was not significant compared to controls (half- time of FRET decay: CTR female vs. FR female; 8.66 min [95% CI: 7.49 min – 9.84 min] vs. 9.66 min [95% CI: 8.92 min – 10.39 min]; t = 1.36; df = 37; p = 0.18; n = 12 CTR females and n = 8 FR females) (Figure 3B-C). In the absence of visual stimulation (darkness), ATP usage was lower than during visual stimulation and unaffected by food restriction, for either sex (Supplemental Figure 1).

ATP usage and orientation selectivity in V1 are more robustly decreased by food restriction in males than in females.

(A) Schemata of ATP imaging experiment. Top: example field of view of V1 layer 2/3 neurons in the ATeam1.03YEMK transgenic mouse (scale bar: 10 μm). Bottom: ATP synthesis inhibitors were used to isolate ATP usage, recorded as a decrease in FRET signal during presentation of natural, outdoor scenes. (B) Normalized ATeam1.03YEMK FRET signal in males and during presentation of natural, outdoor scenes. ATP synthesis inhibitors (arrow) were added to isolate ATP usage. Inset: Time to 50% decay of the ATeam1.03YEMK FRET signal (two-way ANOVA: t = 2.87; df = 37; p = 0. 0067; n = 11 CTR males and 10 FR males). (C) As in (B) but for females (inset: two-way ANOVA: t = 1.36; df = 37; p = 0.18; n = 12 CTR females and n = 8 FR females). (D) Top: Schema of two-photon imaging and sample field of view of V1 layer 2/3 neurons expressing GCaMP6s (scale bar: 10 μm). Bottom: Sample fluorescent signals (grey) from a selected neuron (black circle) in response to 6 drifting gratings of varying orientations. Trial averages are in black. Horizontal bar denotes 2s grating presentation. (E) Mean orientation tuning curves normalized to the response to the preferred orientation for CTR and FR males (data from 13). Note that -90° and +90° conditions correspond to the same visual stimulus. Inset: Orientation selectivity index (OSI; two-way ANOVA: t = 3.75; df = 27; p = 0.0009 n = 8 CTR and 8 FR males). (F) As in (E) but for females (Inset: two-way ANOVA: t = 1.90; df = 27; p = 0.08; n = 7 CTR females and 9 FR females). **p < 0.01; ***p < 0.001; n.s. = not significant. Error bars are S.E.M. Data from male mice are from a previously published dataset obtained under similar conditions13.

Altogether, these findings demonstrate that food-restriction significantly impacts V1 energy usage during visual stimulation in males but not in females.

Orientation selectivity in V1 is significantly reduced under food restriction in males but not in females

We next examined how food restriction impacts visual cortical function. In male mice, reductions in energy usage in V1 during food restriction are associated with a decreased visual coding precision as evidenced by a reduction of orientation selectivity13. Given the non-significant impact of food- restriction on energy usage in female mice, we reasoned that orientation selectivity would also be less impacted. We analysed two-photon calcium imaging data of V1 GCaMP6s-labeled layer 2/3 neurons in female mice viewing drifting gratings (Figure 3D-F; n = 7 CTR male, n = 8 FR male, n = 7 CTR female, and n = 9 FR female mice). Data obtained in females were compared to our previously published datasets obtained under similar conditions in male mice13. Consistent with our hypothesis, in contrast to males in which orientation selectivity was significantly reduced by 27% (OSI: CTR male vs. FR male; 0.58 [95% CI: 0.51 – 0.65] vs. 0.42 [95% CI: 0.36 – 0.49]; t = 3.75; df = 27; p = 0.0009), orientation selectivity was modestly (13%) and not significantly reduced by food-restriction in females (OSI: CTR female vs. FR female; 0.60 [95% CI: 0.52 – 0.67] vs. 0.52 [95% CI: 0.45 – 0.58]; t = 1.90; df = 27; p = 0.08) (Figure 3E-F). Direction tuning, as measured by the direction selectivity index (DSI) was also significantly affected by food-restriction in males, but not in females (Supplemental Figure 3; Males: CTR vs. FR; t = 2.16; p = 0.023; df = 12; n = 6 CTR males and 8 FR males; Females: CTR vs. FR; t = 1.49; p = 0.16; df = 14; n = 7 CTR females and 9 CTR females).

Altogether, our results show that energy usage and coding precision in the visual cortex are largely maintained in female mice under food restriction while they are robustly reduced in males. Such findings cannot be readily attributed to differences in behavioral or attentional state. All animals were imaged whilst in a cardboard tube, and therefore not locomoting. Moreover, we found no difference in pupil size during the presentation of visual stimuli (Supplemental Figure 2).

Bayes factor analysis reveals statistically robust impact of metabolic stress on V1 energy usage and visual coding in males, but not females

We used Bayes factor hypothesis testing to better assess the statistical impact of food restriction on visual cortical function and energy usage in male and female mice (Supplemental Figure 3). Bayes factor analysis uses a Bayesian framework to quantify how much more likely the data is under the alternative hypothesis (i.e. there is an effect of food restriction) versus the null hypothesis (i.e. there is no effect of food restriction), quantified by the Bayes factor (BF10)35,36. A BF10 > 3, for example, means that the data are >3 times more likely under the alternative hypothesis than the null hypothesis. Critically, the BF10 allows for a more informative comparison of the effects of food restriction between sexes, as opposed to comparing p-values, which can be misleading35,36. Across the parameters of energy usage and visual coding that we tested (Figures 1-3), BF10 for males was consistently several-fold greater than that for females. Thus, visual cortical function and energy usage was more robustly modulated by food restriction in males than in females.

Our findings however cannot rule out a potential, smaller impact of food restriction on cortical function and energy usage in females, below the threshold for statistical detection in this study. In particular, for females, we found only 3 out of the 10 parameters had BF10 < 0.33, i.e. the data are 3 times more likely under the null hypothesis than the alternative hypothesis supporting the absence of an effect 35,36(Supplemental Figure 3). This suggests that food restriction may have a modest impact on V1 energy usage or function in females, but below the limits of statistical detection in our study.

Altogether, our results show that energy usage and coding precision in the visual cortex are largely maintained in female mice under food restriction while they are robustly reduced in males.

Discussion

Here, we find that in female mice, energy usage and visual coding in V1 is largely unchanged under food restriction. Using molecular analysis and RNA sequencing, we found that energy-regulating pathways were not significantly changed with food restriction in females, in contrast to males; these included pathways associated with AMPK, mTOR, and PPARα signalling, as well as oxidative phosphorylation. Consistent with this, we found that food restriction in females did not impact visual cortical ATP usage nor orientation selectivity, both of which were reduced by food restriction in males. Our findings suggest that the contribution of neocortex to energy-saving adaptations during food restriction is sex-selective and more pronounced in males; females may rely on other energy-saving strategies, such as reducing reproductive function, to cope with food restriction.

Limitations of the study

Our study focused on the effects of a moderate food restriction protocol over 2-3 weeks on neocortical function and energy usage. This protocol is widely used in metabolic studies investigating the impact of caloric restriction on peripheral physiology, as well as in neuroscience studies to motivate rodents to perform specific behaviors6,15–17. However, it remains unclear to what extent our findings generalize to other protocols of food restriction. Indeed, the duration, magnitude, and pattern of food restriction, as well as macronutrient composition are known to impact the physiological responses to dietary manipulations22,37–39. Moreover, whilst we focussed our study on the adult visual cortex, it remains unclear if the brain shows similar sex-specific energy-saving adaptations in other cortical areas, and how such adaptations change with age6,40.

A key strength of our study is that we consistently found, across a number of experimental methods, that changes in cortical function and energy usage were significant in males, but not in females. However, a critical limitation of our study is that we did not have the statistical power to directly compare the effect size of food restriction between the sexes (e.g. ANOVA interaction effects); to do so would require a substantial increase in experimental group sizes by several fold. We addressed this limitation, in part, using Bayes factor analysis, which revealed that the impact of food restriction in males was indeed more robust than in females. This analysis also revealed that our study did not have sufficiently strong evidence to definitively rule out a smaller effect of food restriction in females, which may not have been statistically detectable.

Role of leptin and sex hormones in food restriction impact on cortical activity

Consistent with past studies, we found that food-restricted female mice resisted weight loss, and therefore required a greater restriction of food to achieve similar weight loss (relative to baseline) as males. In addition, for the same relative weight loss as males, females exhibited minimal reductions in serum leptin levels (secreted by adipocytes), consistent with past studies showing that females resist fat loss during food restriction1–3,5,6. Similar effects have been observed across several mammalian species, including humans, and is thought to reflect the differential importance of fat stores for reproductive success in males and females1–5. Notably, we previously demonstrated that reductions in leptin levels during food restriction are necessary for driving energy-saving changes in visual cortical coding in male mice; leptin therefore acts as a critical signal linking the neocortex to fat stores13. That leptin levels are not significantly affected by food restriction in females may explain the minimal impact food restriction has on cortical energy usage and coding precision in female mice.

In addition to leptin, sex hormones likely contribute to sex-specific adaptations to food restriction41. As with leptin, estrogen suppresses food intake and promotes energy expenditure30, in part by augmenting leptin signalling by increasing leptin-induced phosphorylation of STAT342. Moreover, estrogen can enhance ATP production by sensitizing insulin signalling, enhancing glycolysis and oxidative phosphorylation, and along with progesterone, upregulating mitochondrial gene expression and function30,43,44. Interestingly, in our study, RNA sequencing revealed that estrogen signalling pathways were significantly regulated in male visual cortex with food-restriction, and were unchanged in females. Estrogen may thus have overlapping effects with those induced by food restriction, thereby diminishing further responses to food restriction in females6. Notably, it was recently shown that sex differences in calorie-restriction’s metabolic effects that are observed in young adult mice, are largely absent in older mice, when females’ estrogen levels have declined6. Further studies will be needed to explore whether the resilience of cortical function found in this study in young cycling adult females is also observed in non-cycling mice with reduced oestrogen levels, such as in aged mice.

Sex-specific regulation of cellular energy-regulating pathways

AMPK and mTOR signalling pathways are canonical regulators of cellular energy usage20,23. We found these pathways, along with cortical ATP usage, to be significantly modulated by food-restriction in male, but not female mice. AMPK is activated during metabolic stress, and saves energy by regulating oxidative phosphorylation24,25, and inhibiting cellular pathways associated with energy usage, such as mTOR signalling20. mTOR signalling, by upregulating protein synthesis, increases ATP usage by promoting neuronal excitability, neurite outgrowth, ion channel expression, and synaptic plasticity23. mTOR signalling is therefore ideally situated to regulate how much ATP is used in neuronal function; its inhibition may underlie ATP savings and the associated loss of coding precision with food-restriction in male mice. Notably, previous studies have found sex- and tissue- dependent regulation of AMPK and mTOR signalling, including in response to food restriction41. For example, in skeletal muscle, mTOR signalling is enhanced by fasting selectively in male, but not female mice41,45. Moreover, metabolic challenge by endurance training induces greater AMPK activity in cardiac tissue in male, as opposed to female mice46. Our study extends the sex-specific regulation of these energy-regulating pathways to the neocortex.

Importance of biological sex on the impact of food restriction on cortical function

Food restriction protocols are widely used to motivate animals in behavioural studies examining cortical function15–17. Our findings suggest that these protocols are likely to have greater impacts on basal neocortical function in males, as compared to females. Consistent with this, a recent study has found that food restriction in juvenile mice strongly impacts cognitive flexibility in learning and decision-making in adulthood, specifically for males but not females47,48. Such sex-specific impacts of food restriction on neocortical function are important to take into consideration in the design and interpretation of such studies.

Diet is known to have a considerable impact on neurological diseases, including epilepsy, Alzheimer’s disease, and other neurodegenerative diseases49,50. Whilst calorie restricted diets have proven to be of benefit in this regard, the sex-dependency of such interventions have not been thoroughly investigated, though on the basis of our findings it is conceivable that males and females may respond differently to dietary interventions. The inclusion of both males and females in biomedical research, and the study of sex differences, is therefore of considerable importance for ensuring efficacious treatments are developed for both sexes.

In conclusion, we find that visual cortical coding and energy usage in females are resistant to food restriction, in contrast to males. Our findings establish that sex-specific adaptations in energy usage, previously described in peripheral tissue under metabolic challenge, extends to the neocortex.